Open archaeology, open source?

### 2023-09-20 analysis/_04-open_archaeology.qmd
Rearrange sections a bit

# Open archaeology <!-- Results I (description of what's there) -->

@@

>>>>

# Open archaeology <!-- Results I: description of open archaeology projects -->

@@

## History

Archaeological software development activity has increased significantly over the years. The following charts portray the cumulative growth of code contributions committed and pushed to GitHub repositories, and the number of GitHub repositories that host archaeological software and resources.

```{r plot-gh-cum}
oarch %>% 
  mutate(
    lumped_category = recode(category,
                             "Specifications etc." = "Documents",
                             "Products" = "Documents",
                             "Guides" = "Documents",
                             "Lists and datasets" = "Documents",
                             .default = category),
    lumped_category = factor(lumped_category, c("Scripts",
                                                "Documents",
                                                "Packages and libraries",
                                                "Standalone software")) ) ->
  oarch_lumped

p_gh_commits_cum <- oarch_lumped %>%
  unnest(gh_commits) %>%
  drop_na(sha) %>%
  mutate(date = floor_date(datetime, "week")) %>%
  group_by(lumped_category, date) %>%
  summarise(n_commits = n(), .groups = "drop_last") %>%
  mutate(cum_commits = cumsum(n_commits)) %>%
  ggplot(aes(date, cum_commits,
             fill = lumped_category,
             colour = lumped_category,
             label = lumped_category)) +
  geom_area() +
  annotate("segment", 
           x = ymd("2008-04-01", tz = "UTC"), 
           xend = ymd("2008-04-01", tz = "UTC"),
           y = 16000, yend = 4000,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", 
           x = ymd("2008-04-01", tz = "UTC"), y = 19000,
           label = "GitHub\nlaunched",
           vjust = 0) +
  scale_x_datetime(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::comma) +
  scale_colour_bright(guide = guide_none()) +
  scale_fill_bright() +
  labs(
    y = "cumulative commits",
    x = NULL, fill = NULL
  ) +
  theme_oarch() +
  theme(legend.position = "top")

# TODO: extend to end
p_gh_repos_cum <- oarch_lumped %>%
  unnest(gh_commits) %>%
  drop_na(sha) %>%
  group_by(repo) %>% 
  filter(datetime == min(datetime)) %>% # Only first commits
  mutate(date = floor_date(datetime, "week")) %>%
  group_by(lumped_category, date) %>%
  summarise(n_commits = n(), .groups = "drop_last") %>%
  mutate(cat_total = sum(n_commits)) %>%
  mutate(cum_commits = cumsum(n_commits)) %>%
  ggplot(aes(date, cum_commits,
             fill = lumped_category,
             colour = lumped_category)) +
  geom_area() +
  annotate("segment",
           x = ymd("2008-04-01", tz = "UTC"),
           xend = ymd("2008-04-01", tz = "UTC"),
           y = 40, yend = 10,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text",
           x = ymd("2008-04-01", tz = "UTC"), y = 50,
           label = "GitHub\nlaunched",
           vjust = 0) +
  scale_x_datetime(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::comma) +
  scale_colour_bright(guide = guide_none()) +
  scale_fill_bright(guide = guide_none()) +
  labs(
    y = "cumulative repositories",
    x = NULL, fill = NULL
  ) +
  theme_oarch()

p_gh_commits_cum / p_gh_repos_cum
```

As we can see, archaeologists have been using git from even before GitHub was launched in 2008. But use of git really began to take off around 2014--2015, when we see an uptick in the rate of growth that has continued ever since.

Around this time we also see that GitHub starts being used to host documents and scripts. This may represent a recognition of GitHub's ability to track things other than code, and a willingness to experiment with version control systems as a medium for disseminating work in an open and somewhat nerdy way.

---

- Compare changing proportions of each category on a year by year basis
- Identify temporal trends in the use of licenses

### 2023-09-20 analysis/_07-conclusion.qmd
Rearrange sections a bit

In this paper we take stock of archaeologists' immense creative capabilities. 
First, we surveyed what kinds of software archaeologists are making and tracked progression over time. 
We showed that open source software development in archaeology has seen a rapid and sustained rise beginning around 2014. 
This is marked by a variety of applications and use cases, including the use of git and GitHub to track and host content other than code. 
Moreover, archaeologists are very involved in broader scripting ecosystems, as is evident through the predominant creation of R packages and Python libraries designed to process the rich variety of archaeological information. 
At the same time, archaeologists also create standalone software for more intensive tasks that require greater access to system resources or that warrant more complex user interfaces than what R and Python IDEs are capable of providing.

>>>>

In this study we have sought to document an emerging community of open source software engineers in archaeology and to assess the extent to which collaboration within this community resembles conventional 'open source' forms. 
We have operationalised 'open-source collaborative practices' as the use of certain features of git and GitHub visible to us in data from the GitHub API. 
This excludes a wide range of other potential practices, most obviously the use of other software (though as we demonstrate in section X.X, the git/GitHub combination is overwhelmingly the most popular), but also collaboration through offline or private channels, or forms of collaboration we don't know about. 
We have not directly interviewed those doing the collaborating, though our conclusions do drawn heavily from our experience as members of that community ourselves. 
Our earliest data is from XXX and our study can say little about collaborative software development in archaeology before this point, though we know there was a significant amount of it (@cite).

These caveats notwithstanding, we have documented that open source software development in archaeology has seen a rapid and sustained rise beginning around 2014. 
This is marked by a variety of applications and use cases, including the use of git and GitHub to track and host content other than code. 
Moreover, archaeologists are very involved in broader scripting ecosystems, as is evident through the predominant creation of R packages and Python libraries designed to process the rich variety of archaeological information. 
At the same time, archaeologists also create standalone software for more intensive tasks that require greater access to system resources or that warrant more complex user interfaces than what R and Python IDEs are capable of providing.

### 2023-09-20 analysis/openarchaeo_collaboration.qmd
Rearrange sections a bit

{{< include _01-introduction.qmd >}}
{{< include _02-background.qmd >}}
{{< include _03-data_methodology.qmd >}}
{{< include _04-what_is_open_archaeology.qmd >}}
{{< include _05-history_of_open_archaeology.qmd >}}
{{< include _06-project_activity.qmd >}}
{{< include _07-network.qmd >}}
{{< include _99-conclusion.qmd >}}

>>>>

{{< include _01-introduction.qmd >}}
{{< include _02-background.qmd >}}
{{< include _03-data_methodology.qmd >}}
{{< include _04-open_archaeology.qmd >}}
{{< include _05-collaboration.qmd >}}
{{< include _06-network.qmd >}}
{{< include _07-conclusion.qmd >}}
